---
title: "Grubman et al data set up for sex stratified differential gene expression analysis"
author: "Stella Belonwu"
date: "03/23/2021"
output: html_document
---

Data acquired from Single nucleus RNA-seq from the Entorhinal Cortex\
Single nucleus RNA-seq from Entorhinal Cortex (http://adsn.ddnetbio.com)\

Notes: Dataset has already been QC'ed\
-The raw expression matrix was composed of 33,694 genes and 14,876 cells.\
-Genes without any counts in any cells were filtered out (needed to have at least 2 transcripts in 10 cells).\
-Discarded 100 PMI-associated genes.\
-Discarded cells outside 5th and 95th percentile of genes detected and unique molecular identifiers.\
-Cells with more than 10% of their UMIs assigned to mitochondrial genes were filtered out.\
-Matrix was normalized with a scaling factor of 10000 before FindVariableGenes was used to define variable genes with the parameters x.low.cutoff = 0.0125, x.high.cutoff = 3, and y.cutoff = 0.5.\
-No need to subset mitochondrial counts, keep below 10% as paper did\
-No need to subset cells, already done by paper\
-Remove pctMT, nGene, nUMI, agrees with Seurat's percent.nt, nFeature_RNA, and nCount_RNA respectively\

Plan:\
-Run regular normalization\
-Batch correct with seurat integration and SCT\
-Subset datasets for DE analysis\

Input data\ 
```{r}
setwd("/your directory/")
grub_Counts<-read.table("scRNA_rawCounts_grubman.tsv",header= TRUE, sep = "\t")
grub_meta<- read.table("scRNA_metadata_grubman.tsv",header= TRUE, sep = "\t")
grub_supp<- read.csv("scRNA_supp_grubman.csv" ,header= TRUE, sep = ",", stringsAsFactors=FALSE)
```

Input packages\
```{r}
library(dittoSeq)
library(ggplot2)
library(Seurat) 
library(dplyr)
library(sctransform)
library(cowplot)
library(magrittr)
```

Prep metadata and count data\  
```{r}
colnames(grub_meta)
grub_meta2<- grub_meta[,c(1:3,7,8)]
colnames(grub_meta2)[4]<- "grub_cellType" #original cell labels

unique(grub_meta2$batch)
unique(grub_meta2$batchCond)
unique(grub_meta2$patient)
grub_meta2$patient<- as.character(grub_meta2$patient)
rownames(grub_meta2)<- grub_meta2$sampleID
 
#Remove unassigned samples and keep all present ApoE groups
colnames(grub_meta2)[c(2,3,5)]<- c("Batch", "Patient","Diagnosis") 
grub_supp2<- grub_supp[c(1:6,9:14),]
grub_supp2$patient<- c("AD1","AD2", "AD3", "AD4" ,"AD5", "AD6" , "Ct1", "Ct2", "Ct3", "Ct4", "Ct5", "Ct6")
grub_supp3<- grub_supp2[!is.na(grub_supp2$ApoE), c(8,2:6)] #AD1 is 34, AD3 is 44 *update from authors

table(grub_supp3$ApoE) #1 24, 5 33, 4 34,  2 44
table(grub_supp3$Diagnosis) #6 AD, 6 Control

sapply(grub_meta2,class)
grub_meta2$Diagnosis<- as.character(grub_meta2$Diagnosis)
grub_meta2$Diagnosis<- ifelse(grub_meta2$Diagnosis == "ct", "Control", grub_meta2$Diagnosis)

colnames(grub_supp3)[c(1,3)]<- c("Patient", "Sex")
unique(grub_supp3$Patient) #"AD1" "AD2" "AD3" "AD4" "AD5" "AD6" "Ct1" "Ct2" "Ct3" "Ct4" "Ct5" "Ct6"
unique(grub_meta2$Patient) #  "AD5"   "AD6"   "AD-un" "AD3"   "AD4"   "Ct6"   "Ct5"   "Ct-un" "Ct4"   "Ct3"   "Ct1"   "Ct2"   "AD2"   "AD1"  

grub_meta3<- grub_meta2[grub_meta2$Patient != "AD-un",  ] #13029  
grub_meta3<- grub_meta3[grub_meta3$Patient != "Ct-un",  ] #12770 

grub_supp3<- grub_supp3[grub_supp3$ApoE != "24",  ] #Remove the APOE24 sample
grub_meta3<- grub_meta3[grub_meta3$Patient %in% grub_supp3$Patient, ] #11704
 
#Match Count Data   
grub_Counts2<- grub_Counts
rownames(grub_Counts2)<- grub_Counts2$geneName
grub_Counts2$geneName<- NULL
dim(grub_Counts2) #10850genes 13214cells
  
grub_meta4<- dplyr::inner_join(grub_meta3, grub_supp3[,c(1:4,6)], by="Patient") 
rownames(grub_meta4)<- grub_meta4$sampleID
grub_meta4$sampleID<- NULL 
grub_meta4$TAG<- rownames(grub_meta4)
sapply(grub_meta4, class)
grub_meta4$ApoE %<>% factor
grub_meta4$Sex %<>% factor
grub_meta4$Batch %<>% factor

grub_Counts3<- grub_Counts2[,colnames(grub_Counts2) %in% rownames(grub_meta4)] #10850 genes, 11704 cells 
```

Create Seurat object\ 
```{r}
grub<-  CreateSeuratObject(counts = grub_Counts3,project = "grub", min.cells = 3, min.features = 200, meta.data = grub_meta4)

# The [[ operator can add columns to object metadata. This is a great place to stash QC stats
grub[["percent.mt"]] <- PercentageFeatureSet(grub, pattern = "^MT-")

range(grub$nFeature_RNA) # 274 1632 
range(grub$percent.mt) #0.000000 9.952324

# Visualize QC metrics as a violin plot
VlnPlot(grub, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)

# Use FeatureScatter to visualize feature-feature relationships i.e. columns in object metadata, PC scores etc.
plot1 <- FeatureScatter(grub, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(grub, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
CombinePlots(plots = list(plot1, plot2))
```

Normalize object\
```{r}
grub<- NormalizeData(object = grub) 
grub <- FindVariableFeatures(object = grub, selection.method = "vst", nfeatures = 2000, x.low.cutoff = 0.0125, x.high.cutoff = 3,  y.cutoff = 0.5) 

top10_varfeatures <- head(VariableFeatures(grub), 10)

plot1 <- VariableFeaturePlot(grub)
plot2 <- LabelPoints(plot = plot1, points = top10_varfeatures, repel = TRUE)
CombinePlots(plots = list(plot1, plot2) ) 
```

ScaleData to center the gene expression & Run PCA, UMAP, TSNE (make test plots)\
```{r}
length(rownames(grub)) #10850 
all.genes <- rownames(grub)  
grub <- ScaleData(object = grub, features = all.genes)

grub <- RunPCA(object = grub, features = VariableFeatures(object = grub))
ElbowPlot(grub, ndims = 50)
grub <- RunUMAP(grub, dims = 1:10)
grub <- RunTSNE(grub, dims = 1:10)

#barplot(1:40, col = dittoColors()[1:40]) #run to select colors
unique(grub$Batch)
dittoDimPlot( object = grub, "Batch",reduction = "pca", color.panel = dittoColors()[c(39,15,7,36,25,17)])
dittoDimPlot( object = grub, "Batch",reduction = "umap" ,color.panel = dittoColors()[c(39,15,7,36,25,17)])
dittoDimPlot( object = grub, "Batch",reduction = "tsne" ,color.panel = dittoColors()[c(39,15,7,36,25,17)])
dittoDimPlot( object = grub, "grub_cellType",reduction = "pca")
dittoDimPlot( object = grub,  "grub_cellType",reduction = "umap")
dittoDimPlot( object = grub,  "grub_cellType",reduction = "tsne")
```

Save pre-integration object\
```{r}
saveRDS(grub, file = "grub_pre-integration.rds")
```
  
Batch Correction: Split, SC Transform, Seurat Integrate\
DO NOT RUN SCALE DATA AFTER INTEGRATION\
https://satijalab.org/seurat/v3.0/sctransform_vignette.html\
https://satijalab.org/seurat/v3.0/integration.html\
fix space error: https://satijalab.org/seurat/v3.0/future_vignette.html\
```{r}
#SCTransform replaces NormalizeData, ScaleData, and FindVariableFeatures; DON'T RUN SCALE DATA AFTER
grub.list <- SplitObject(grub, split.by = "Batch")

for (i in 1:length(grub.list)) {
    grub.list[[i]] <- SCTransform(grub.list[[i]], verbose = TRUE) 
}
 
#Select features for downstream integration, and run PrepSCTIntegration, which ensures that all necessary Pearson residuals have been calculated.
grub.features <- SelectIntegrationFeatures(object.list = grub.list, nfeatures = 3000)
options(future.globals.maxSize = 1000 * 1024^3) #Give at least 1GB
grub.list <- PrepSCTIntegration(object.list = grub.list, anchor.features = grub.features, verbose = FALSE)

#Identify anchors and integrate the datasets. Be sure to set normalization.method = 'SCT':
grub.anchors <- FindIntegrationAnchors(object.list = grub.list, normalization.method = "SCT", anchor.features = grub.features, verbose = FALSE)
grub.integrated <- IntegrateData(anchorset =grub.anchors, normalization.method = "SCT", verbose = FALSE)

#Adding a command log without an assay associated with it -> not an issue (https://github.com/satijalab/seurat/issues/2505)
grub.integrated@assays$integrated@counts
```

Moving forward, use "RNA" for DE analysis and "integrated" for dimensionality reduction\
```{r} 
DefaultAssay(grub.integrated) <- "integrated" # switch to integrated assay. The variable features of this assay are automatically set during IntegrateData

# Run the standard workflow for visualization and clustering
grub.integrated <- RunPCA(grub.integrated, verbose = FALSE)
ElbowPlot(grub.integrated, ndims =50)
grub.integrated <- RunUMAP(grub.integrated, dims = 1:10)
grub.integrated <- RunTSNE(grub.integrated, dims = 1:10)
 
#QC plots
p1 <- DimPlot(grub.integrated, reduction = "umap", group.by = "Batch")
p2 <- DimPlot(grub.integrated, reduction = "umap", group.by = "Diagnosis")
plot_grid(p1, p2) 
VlnPlot(grub.integrated, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
```

Cell Identity analysis using human brain markers from BRETIGEA\
Calculate a module score for each cell type using Seurat’s AddModuleScore and the top 200 genes for each cell type\
Observe expression of Marker Genes and UMAP cell identity homogeneity\
Assign cell based on the highest score in each\
Identify hybrid cells as cells where the top two highest cell type scores are similar (within 20%) of each other\
Observe expression of Marker Genes and UMAP cell identity homogeneity\
```{r}
library(BRETIGEA)
markers_df_human_brain
dim(markers_df_human_brain) #5500,2 
table(markers_df_human_brain$cell) #1000 each: ast,end, mic, neu,oli,opc (500)

brain_markers<- markers_df_human_brain #5500
brain_markers<- brain_markers[brain_markers$markers %in% all.genes,] #3007
table(brain_markers$cell) #ast 616, end 394, mic 428, neu 690, oli 650, opc 229
length(unique(brain_markers$markers)) #2956

library(dplyr)
brain_markers2<-  brain_markers  %>%
  group_by(cell) %>%
  slice(1:200)
table(brain_markers2$cell)  

neu_markers<- list(brain_markers2$markers[brain_markers2$cell == "neu"]) 
end_markers<- list(brain_markers2$markers[brain_markers2$cell == "end"]) 
ast_markers<- list(brain_markers2$markers[brain_markers2$cell == "ast"]) 
mic_markers<- list(brain_markers2$markers[brain_markers2$cell == "mic"]) 
oli_markers<- list(brain_markers2$markers[brain_markers2$cell == "oli"]) 
opc_markers<- list(brain_markers2$markers[brain_markers2$cell == "opc"]) 

grub2<- grub.integrated
DefaultAssay(grub2) <- "SCT" 

grub2<-AddModuleScore(object=grub2, features =  neu_markers, pool = NULL,  nbin = 24,  ctrl = 100,  k = FALSE,  assay = NULL, name = 'neu',  seed = 1,  search = FALSE)

grub2<-AddModuleScore(object=grub2, features =  end_markers,  pool = NULL,  nbin = 24,  ctrl = 100,  k = FALSE,  assay = NULL,  name = 'end',  seed = 1,  search = FALSE)

grub2<-AddModuleScore(object=grub2,  features = ast_markers,  pool = NULL,  nbin = 24,  ctrl = 100,  k = FALSE,  assay = NULL,  name = 'ast',  seed = 1,  search = FALSE)

grub2<-AddModuleScore(object=grub2, features =  mic_markers,  pool = NULL,  nbin = 24,  ctrl = 100,  k = FALSE,  assay = NULL,  name = 'mic',  seed = 1,  search = FALSE)

grub2<-AddModuleScore(object=grub2, features =  oli_markers,  pool = NULL,  nbin = 24,  ctrl = 100,  k = FALSE,  assay = NULL,  name = 'oli',  seed = 1,  search = FALSE)

grub2<-AddModuleScore(object=grub2, features =  opc_markers,  pool = NULL,  nbin = 24,  ctrl = 100,  k = FALSE,  assay = NULL,  name = 'opc',  seed = 1,  search = FALSE)
```

Defined a cell as a hybrid cell if the difference between the first and second highest cell type scores are within 20% of the highest cell type score: (x1 − x2)/x1 < 0.2 x1=highest, x2=second highest\
```{r} 
dittoDimPlot( object = grub2, "neu1",reduction = "umap")
dittoDimPlot( object = grub2, "end1",reduction = "umap")
dittoDimPlot( object = grub2, "ast1",reduction = "umap")
dittoDimPlot( object = grub2, "mic1",reduction = "umap")
dittoDimPlot( object = grub2, "oli1",reduction = "umap")
dittoDimPlot( object = grub2, "opc1",reduction = "umap")

#Make cell type column
grub_new_meta<- grub2@meta.data
grub_new_meta$cell_type<- names(grub_new_meta[, 24:29])[apply(grub_new_meta[, 24:29],1,which.max)]
table(grub_new_meta$cell_type)
#ast1 end1 mic1 neu1 oli1 opc1 
#2144  106  453  666 7373  962 

#Max and second max
grub_new_meta$x1<- apply(grub_new_meta[, 24:29], 1, max)
grub_new_meta$x2<- apply(grub_new_meta[, 24:29], 1, function(x) x[order(x)[5]]) #in decreasing order
grub_new_meta$x1_x2<- (grub_new_meta$x1 - grub_new_meta$x2)/grub_new_meta$x1  
grub_new_meta$x1_x2<- ifelse(grub_new_meta$x1_x2 < 0.2, 1,0) #if the difference between top two scores is less than 20%, 1, if not 0
table(grub_new_meta$x1_x2) #11385 0, 319 1

#Make second column including hybrid
grub_new_meta$cell_type2<- grub_new_meta$cell_type
grub_new_meta$cell_type2<- ifelse(grub_new_meta$x1_x2 == 1, "hybrid", grub_new_meta$cell_type2 )
table(grub_new_meta$cell_type2) 
#ast1   end1 hybrid   mic1   neu1   oli1   opc1 
#2105    101    319    434    619   7203    923 

#remove numbers from cell type name
grub_new_meta$cell_type<- gsub("1", "", grub_new_meta$cell_type)
unique(grub_new_meta$cell_type)

grub_new_meta$cell_type2<- gsub("1", "", grub_new_meta$cell_type2)
unique(grub_new_meta$cell_type2)

write.csv(grub_new_meta, file="2020_09_13_grub_meta_w_cell_id_scores.csv")
#remove unnecessary columns
colnames(grub_new_meta)
grub_new_meta2<- grub_new_meta[,c(1:15,30,34)]
colnames(grub_new_meta2)

#replace metadata
grub2@meta.data<- grub_new_meta2
```

Compare cell type identification methods\
original paper, prehybrid & post hybrid\
```{r}
table(grub_new_meta2$grub_cellType, grub_new_meta2$cell_type)
table(grub_new_meta2$grub_cellType, grub_new_meta2$cell_type2)

#rearrange color scheme to properly compare cell type IDs
unique(grub2$grub_cellType) #"oligo"   "astro"   "endo"    "mg"      "neuron"  "OPC"     ["unID"]    ["doublet"]
unique(grub2$cell_type) #"oli" "ast" "opc" "neu" "end" "mic"
unique(grub2$cell_type2) #oli"    "ast"    "opc"    "neu"    "end"    "mic"    ["hybrid"]

dittoColors()
dittoDimPlot( 'grub_cellType', object = grub2, reduction = "umap", color.panel = c("#E69F00" ,"#dd1c77","#56B4E9" ,"#009E73", "#F0E442","#D55E00", "#0072B2","#666666" ))
dittoDimPlot( 'cell_type', object = grub2, reduction = "umap", color.panel = c("#E69F00" ,"#56B4E9" ,"#009E73", "#F0E442", "#D55E00" , "#0072B2"))
dittoDimPlot( 'cell_type2', object = grub2, reduction = "umap", color.panel = c("#E69F00" ,"#56B4E9" ,"#dd1c77" ,"#009E73", "#F0E442",  "#D55E00", "#0072B2"))
```

Save Seurat object with and without hybrids\
```{r}
setwd("/your directory/")
saveRDS(grub2, file = "grub_corrected_full_hybrid.rds")

grub3 <- subset(grub2, subset = cell_type2 != "hybrid")
grub2 #11704 samples
grub3  #11385 samples
saveRDS(grub3, file = "grub_corrected_full_nohybrid.rds")
```